\section{Introduction}\label{sec:intro}

Type-based reasoning has become a standard approach to functional program verification.
Refinement types are a prominent example: they support overapproximate safety reasoning by embedding logical safety properties directly into types.
For a typing judgment $\Gamma \vdash e : \tau$, the result of $e$ may lie in the set described by $\tau$ under that context.
Not every behavior admitted by $\tau$ needs to be realized by $e$, so $\tau$ is an overapproximation of $e$'s behavior.
This relation must hold for all contexts admitted by $\Gamma$, yielding the usual safety guarantee: a well-typed program never goes wrong under any \emph{demonic} context.

Dually, we can adopt an underapproximate interpretation, where $\tau$ gives a lower bound of $e$'s behavior under an angelic style context.
This view has proved useful in incorrectness reasoning, which aims to show that unexpected behaviors must occur~\cite{OHP19}.
For example, Incorrectness Logic~\cite{OHP19} uses underapproximate triples $[P]\;\Code{C}\;[Q]$, where the postcondition $Q$ is treated as a lower bound of $\Code{C}$'s behavior under an angelic precondition $P$.
% \begin{align*}
%     &\overtext{\forall \sigma'. Q(\sigma')} \impl \undertext{\exists \sigma. P(\sigma) \land (\sigma, \Code{C}) \Downarrow \sigma' }
% \end{align*}\noindent
The states in $Q$ underapproximate the states reachable from some state in $P$ by executing $\Code{C}$.
Importantly, $P$ is interpreted angelically: the context can choose a suitable pre-state $\sigma$ to justify a target post-state $\sigma'$.
This style of reasoning is widely studied in program logic, especially to reduce overwhelming false positives in bug finding and reachability analyses.

However, analogous incorrectness reasoning is not yet established in type-based verification.
Coverage Type~\cite{CoverageType} interprets types as underapproximations of test-input-generator behavior, and can prove coverage completeness (e.g., all binary search trees can be generated nondeterministically).
But it does not provide the angelic-style context central to Incorrectness Logic.
Meanwhile, refinement-typing workflows are often extended with external mechanisms such as symbolic execution to search for concrete counterexamples, instead of integrating underapproximation reasoning directly into typing judgments~\cite{RefinementTypeSymbolicExecution}.
This split leaves overapproximate typing judgments and underapproximate witnesses at different methodological layers, limiting the modularity and reuse that type-system-native reasoning would otherwise provide.

\begin{figure}[h!]
    \centering
\begin{minted}[xleftmargin=5pt, numbersep=4pt, linenos = true, fontsize = \footnotesize, escapeinside=??]{OCaml}
    let rec merge (l1: elem list) (l2: elem list) : elem list =
        match l1, l2 with
        | l1, [] -> l1
        | [], l2 -> l2
        | h1 :: t1, h2 :: t2 ->
            let t = merge t1 t2 in (* t contains elements in t1 *)
            if h1 < h2 then h1 :: h2 :: t (* h2 may be greater than elements in t1 *)
            else if h1 > h2 then h2 :: h1 :: t
            else h1 :: t
\end{minted}
\caption{A function that merges two sorted lists. If either of
the input lists are empty, \Code{merge} returns the other (lines
$3-4$). Otherwise, the function merges the tails of the two inputs and appends the two top elements ahead of the merged result. }
\label{fig:merge}
\end{figure}


To illustrate how angelic context helps both correctness and incorrectness reasoning, consider \autoref{fig:merge}.
The function \Code{merge} is a core component of merge sort, and should preserve sortedness of the output list.
However, verification fails because the implementation is buggy:
\begin{align*}
    \texttt{Counterexample:}&\quad  \text{ when } \Code{l1} = [1;2] \text{ and } \Code{l2} = [3;4] \text{ then } \mstep{\Code{merge}\;[1;2]\;[3;4]}{[1;3;2;4]}
\end{align*}\noindent
The bug occurs because the code assumes that after comparing $\Code{h1}$ and $\Code{h2}$, $\Code{h2}$ is always the second-smallest element (line $7$).
Even when $\Code{h1} < \Code{h2}$, $\Code{h2}$ may still be greater than some elements in $\Code{t1}$, which can break sortedness.
However, the counterexample above does not answer the following questions:
\begin{align*}
    \texttt{Question:}&\quad \text{ can } \Code{merge}\;[1]\;\Code{l2} \text{ be erroneous?} \quad \judgfail
    \\\texttt{Question:}&\quad \text{ can } \Code{merge}\;[1;2;3]\;\Code{l2} \text{ be erroneous?} \quad \judgok
\end{align*}\noindent
In fact, when the first list is empty or a singleton, $\Code{merge}$ behaves correctly.
These questions characterize ``how wrong'' the function is, which helps localize the bug.
The function is correct for the first two $\Code{match}$ branches (lines $3$--$4$), and also for the case $\Code{h1}=\Code{h2}$ (line $7$).
This perspective also avoids false-positive alarms on correct runs, e.g., $\Code{merge}\;[1;2]\;[0;4] = [0;1;2;4]$.
Since listing potentially infinitely many counterexamples is not practical, we ask the following mixed-modality incorrectness question:

\textbf{Mixed-modality context:} \emph{For every sorted list $\Code{l1}$, does there exist another sorted list $\Code{l2}$ such that the merged list is unsorted?}

\noindent This specification combines a demonic part and an angelic part.
The first list $\Code{l1}$ is chosen demonically, as in standard refinement typing; then the context can choose $\Code{l2}$ angelically to witness unsorted output.
For instance, given $\Code{l1}=[1;2]$, it can pick $\Code{l2}=[3;4]$ (which exposes the bug), rather than $\Code{l2}=[0;4]$ (which does not).

There is a deep correspondence between underapproximation (resp.\ overapproximation) and angelic (resp.\ demonic) context.
Consider an angelic context $\Gamma_x$ that can assign arbitrary natural numbers to $x$ as needed.
Then the term $x+1$ has a natural lower-bound interpretation over positive numbers.
For example, to witness that $x+1$ can evaluate to a positive number $n$, the context can choose $x \mapsto n-1$.
This reverse-style reasoning---deciding inputs from desired outputs---is the essence of incorrectness reasoning.
Notably, $x+1$ is deterministic; yet it still admits meaningful lower-bound reasoning, as in prior underapproximate type-based work targeting random test-input generators~\cite{CoverageType}.

Enabling mixed-modality context is more challenging than in underapproximate program logics, which usually model context as a single global state with a uniform demonic or angelic interpretation.
In functional languages, context is curried: a function can specify context requirements parameter by parameter, enabling finer-grained mixed-modality structure.
Hence we must distinguish whether two context-provided parameters are fully disjoint.
For example, consider a context $\Gamma_{xy}$ that can assign values to $x$ and $y$ angelically, but only the same value to both.
This context remains angelic for each variable in isolation, yet under $\Gamma_{xy}$ the term $x-y$ can only be zero.
Higher-order functions further complicate dependencies, because the context may provide a function whose own behavior depends on additional contextual choices.


% Unifying over- and underapproximate reasoning within one semantic framework is not just a theoretically elegant proposal; it has been a recurring goal in verification theory and tooling.
% Classic compositional may/must analysis already illustrates the potential synergy: may-style reasoning prunes infeasible regions, while must-style evidence narrows witnesses and localizes specification or implementation defects~\cite{CompositionalMayMust}.
% In current type-based verification practice, however, overapproximate and underapproximate reasoning are still often separated at the judgment level.
% For example, refinement-typing workflows often rely on external mechanisms such as symbolic execution to search for concrete counterexamples to refinement obligations, rather than integrating underapproximate reasoning directly into typing judgments~\cite{RefinementTypeSymbolicExecution}.
% This split leaves overapproximate typing judgments and underapproximate witnesses at different methodological layers, limiting the modularity and reuse that type-system-native reasoning would otherwise provide.
% Our goal is therefore not to bolt must-style analysis onto refinement typing, but to internalize both modalities uniformly in typing rules for higher-order functional programs.

To address these challenges, we treat context as \emph{resources} and introduce a \emph{resource algebra} for principled, compositional decomposition and combination of contextual pieces.
Resources $r_x$ and $r_y$ may combine dependently (as in $\Gamma_{xy}$) or disjointly as $r_x \times r_y$.
In the disjoint case, demonic/angelic decisions in $r_x$ are independent from those in $r_y$.
Therefore, for $x-y$ under $r_x \times r_y$, we can choose $x$ and $y$ independently and realize any natural-number result (e.g., set $y=0$ and vary $x$ arbitrarily).
From this resource-based view, we introduce a \emph{context logic} based on bunched implication.
It treats overapproximation and underapproximation as first-class modalities, and supports fine-grained mixing both dependently (via conjunction $\chland$) and disjointly (via separating composition $\chstar$).
We further design \emph{context types}, a bunched modal type system interpreted in this logic, to track complex data-flow dependencies in tree-structured contexts and support framing-based local reasoning.
Functions can require disjoint or dependent input resources through $\ctwand$ and $\ctsarr$, respectively, enabling context-sensitive reasoning during application.


\paragraph{Contributions.}
We summarize the main technical contributions as follows.
\begin{itemize}[leftmargin=*]
\item We interpret context in functional languages as a resource, and introduce a resource algebra for compositional decomposition and combination of contextual pieces.
\item We define a unified modal \emph{context logic} that supports dependent and disjoint mixing of overapproximation/demonic and underapproximation/angelic modalities.
\item We design \emph{context types}, a bunched modal type system interpreted in our context logic, to support both correctness and incorrectness reasoning for even deterministic functional programs. We later show this type system also works for nondeterministic programs.
\item We provide multiple examples and a case study showing how the system simulates existing reasoning styles and how mixed-modality context helps localize buggy program components.
\end{itemize}

% This paper develops \emph{resolution types}, a resource-modal refinement type system that keeps over- and underapproximation as first-class modalities and allows fine-grained mixing of the two in arbitrary ways.
% For a nondeterministic program, demonic execution corresponds to overapproximation (all possible executions must be as expected), while angelic execution corresponds to underapproximation (there exists some execution as expected).
% For example, given two reductions $\mstep{e}{v_1}$ and $\mstep{e}{v_2}$, a demonic resolution considers all combinations of these reductions (to $v_1$, to $v_2$, or both), while an angelic resolution selects as needed.
% Moreover, the resolving power that enables demonic or angelic choices is treated as a resource in the type system.
% At the extreme, for a singleton-valued term (e.g., $\mstep{1+1}{2}$), angelic and demonic resolutions collapse, since there is no space for them to differ.
% With this resource-based view, we build an underlying resource algebra and resource logic that provide a Kripke semantics for over- and underapproximate modalities.

% This resource-based view also addresses both dependency tracking and function duplicability.
% In general, over- and underapproximate modalities can be related in two ways: dependent or independent, which naturally recalls classical bunched typing~\cite{BunchedType} and other resource-based disjointness reasoning approaches.
% For example, our type system provides a \emph{magic wand} ($\ctwand$), in addition to the normal arrow ($\ctsarr$), to express independence requirements between function parameters:
% \begin{align*}
%     \vdash \Code{existsElem} : n{:}\ort{\Nat}{\top} \ctwand \urt{\List[\Nat]}{\top} \ctwand \urt{\Bool}{\vnu = \true} 
% \end{align*}\noindent
% With bunched typing, we can track complex data-flow dependencies in a tree-structured context and apply framing rules for local reasoning.
% To express function duplicability, we also introduce a \emph{persistent} modality $\ctpers$, meaning the function is independent of demonic or angelic decisions made by the context (i.e., the contextual resource is singleton).
% The function then makes decisions solely from the nondeterminism encapsulated in its body, and is therefore allowed to produce different results across multiple applications.



% Semantically, we reuse the alternating structure familiar from semantic games and from quantifier alternation in QBF:
% some unknowns are modeled as externally fixed resolutions subject to \emph{may}-style overapproximate bounds,
% others as verifier-guided choices drawn from \emph{must}-style underapproximate witness sets that may shrink as derivations proceed.
% Following standard terminology, these correspond to universal/demonic and existential/angelic choice, respectively:
% underapproximate assumptions may still be strengthened by narrowing witnesses, whereas overapproximate assumptions are interpreted may-style and are not arbitrarily refinable during verification without additional constraints.
% Typings record proof-relevant \emph{strategy information} on hypotheses; under this reading, a typing derivation can be viewed as a strategy in which over-refinements constrain universal choices and under-refinements constrain existential ones.
% For the mixed specification above, the universally quantified tag \(n\) is fixed first; the existentially controlled part must still realize the stated coverage obligation at that \(n\).

% We treat strategy as a resource in the type system.
% Operationally, an underapproximated refinement type in the context indicates that the prover is free to make a choice \emph{linearly}; once fixed, it cannot be renegotiated.
% The resource reading also gives a modular way to track dependencies between refinements.
% Instead of encoding dependencies solely through explicit reachability-style dependency annotations on
% types~\cite{ReachabilityTypes}, we structure contexts with additive \(\ctcomma\) and separating \(\ctstar\), following separation logic~\cite{SeparationLogic} and bunched typing~\cite{BunchedType}.
% The resulting bunched contexts distinguish which refinements may share witness state and which must be treated as disjoint, so that standard local reasoning principles (including framing) apply.



%% ZZ: not sure if we need to highlight this part
% Following~\autoref{sec:overview}, those phenomena are modelled through \emph{strategic contexts} built
% from nonempty compatible substitution sets (\emph{strategy resources}): separating products isolate
% independent strategic fragments while additive structure records branching-induced correlation and
% type-level sums keep control-flow splits explicit.
% In read/write terms, disjoint fragments behave like isolated ``write scopes'' for underapproximate slack,
% while overbounded fragments mimic duplicable read-only permissions unless the typing rules force explicit
% correlation.
