\section{Supporting Nondeterminism}\label{sec:extension}

The core language of Section~\ref{sec:context-typing} is
deterministic: every term reduces to at most one value under any
environment.  This section extends the language with
\emph{nondeterminism}, expressed in terms of primitive generators that
can produce multiple values across different executions, and
identifies the adjustments required to the type denotation.
Concretely, our language now supports generators over primitive types:
\begin{align*}
  \primop ::= \quad & \ldots ~|~ \Code{bool\_gen()} ~|~
  \Code{nat\_gen()} ~|~ \ldots
\end{align*}
The nondeterministic choice $e_1 \oplus e_2$ is then expressible as
$\zif{\Code{bool\_gen()}}{e_1}{e_2}$.  These generators serve as the
``impure external environments'' introduced informally in
Section~\ref{sec:context-typing}: the environment $\Envir_x(e) \doteq
\zlet{x}{\Code{nat\_gen()}}{e}$ is now a first-class term that
introduces a nondeterministic binding for $x$ whose capability is
\emph{independent} of the surrounding context at each invocation.

Nondeterminism raises two technical challenges.  First, nondeterministic
generators can be invoked multiple times, producing independent
capabilities at each call; the type system must account for this
\emph{resource duplication}.  Second, nondeterminism breaks the
single-valued assumption underlying our type denotation, requiring a
generalization of the denotation of function and sum types.
% (\S\ref{subsec:nondet-denotation}).  We conclude with case studies
% (\S\ref{subsec:case-studies}) demonstrating the expressiveness of the
% resulting system.

\subsection{Resource Duplication and Persistence}\label{subsec:duplication}

In the deterministic core language, disjoint capabilities are \emph{external only}:
evaluation does not introduce new capabilities.
For example, the only inhabitant of $\ort{\Unit}{\top} \ctsarr \urt{\Nat}{\top}$ should be function that returns the captured angelic binding $z$:
\begin{align*}
  z{:}\urt{\Unit}{\top} \vdash \zzlam{x}{z} : \ort{\Unit}{\top} \ctsarr \urt{\Nat}{\top}
\end{align*}\noindent
Two successive calls to $\zzlam{x}{z}$ must therefore yield dependent results even with disjoint arguments:
\begin{align*}
  z{:}\urt{\Unit}{\top} \vdash \zlet{f}{\zzlam{x}{z}}{\zlet{y_1}{f\;()}{\zlet{y_2}{f\;()}{e}}} : \tau
\end{align*}\noindent
This is guaranteed by our bunched typing rules \textsc{T-Let} and \textsc{T-LetD}, which enforce the ``dependent xor consumed'' policy, where the rule \textsc{T-Let} allows the function $f$ to be used in both the binding expression and the body of a \Code{let}-expression, but the application results can only be dependent.
\begin{align*}
  &\underline{\Gamma \ctcomma (z{:}\urt{\Unit}{\top} \ctcomma f{:}\ort{\Unit}{\top} \ctsarr \urt{\Nat}{\top}) \ctcomma y_1{:}\urt{\Nat}{\top} \vdash \zlet{y_2}{f\;()}{e} : \tau} \tag{\textsc{T-Let}}
  \\&\Gamma \ctcomma (z{:}\urt{\Unit}{\top} \ctcomma f{:}\ort{\Unit}{\top} \ctsarr \urt{\Nat}{\top}) \vdash \zlet{y_1}{f\;()}{\zlet{y_2}{f\;()}{e}} : \tau
\end{align*}\noindent
On the other hand, the rule \textsc{T-LetD} allows producing disjoint results but disallows using the function $f$ in the body of a \Code{let}-expression. As shown below, the binding $f$ is consumed from the type context, so the second application cannot be typed.
\begin{align*}
  &\underline{\Gamma \ctstar y_1{:}\urt{\Nat}{\top} \vdash \zlet{y_2}{f\;()}{e} : \tau \quad \judgfail}  \tag{\textsc{T-LetD}}
  \\&\Gamma \ctstar (z{:}\urt{\Unit}{\top} \ctcomma f{:}\ort{\Unit}{\top} \ctsarr \urt{\Nat}{\top}) \vdash \zlet{y_1}{f\;()}{\zlet{y_2}{f\;()}{e}} : \tau
\end{align*}\noindent

With a nondeterministic generator like $\Code{nat\_gen()}$, however,
successive calls produce genuinely independent values. 
The key
observation is that nondeterministic generators like
$\Code{nat\_gen()}$ must be given a \emph{persistent} type:
\begin{align*}
  \Code{nat\_gen} : \ctpers (\ort{\Unit}{\top} \ctsarr \urt{\Nat}{\top})
\end{align*}
Persistence allows a binding in the type context to be duplicated, i.e.,
\begin{align*}
  x{:}\ctpers \tau \; \ctsubseteq{\{x\}} \; x{:}\ctpers \tau \ctstar x{:}\ctpers \tau 
\end{align*}\noindent
With the help of the rule \textsc{T-LetD}, the resource can be freely duplicated across multiple function applications without entangling the corresponding
capabilities.

\paragraph{Duplicable resource and persistent modality} As in other BI-logics that support the persistent modality~\cite{Iris}, the persistent modality is defined based on the \emph{duplicable} resource with respect to the product operation in the BI-algebra. The duplicable resource is idempotent with respect to the product operation, i.e., $r \times r = r$, which allows us to duplicate the generator types in the type context.
In our setting, $r \times r = r$ happens only when $r$ is a \emph{singleton} capability that contains only one environment. Following this idea, we can define the persistent modality in our context logic, which means a formula does not rely on the different choices provided by the capability, i.e., the capability is singleton.

\begin{definition}[Persistent modality] We introduce the standard persistence modality $\chpers P$ in our context logic such that:
  \begin{align*}
    r \models \chpers P \text{ iff } \exists \sigma. \{\sigma\} \sqsubseteq r \land \{\sigma\} \models P
  \end{align*}\noindent
  Note that we do not require $r$ to be singleton over its entire variable domain, which would violate Kripke monotonicity: if a capability $r$ is singleton, then for $r'$ such that $r' \sqsubseteq r$, $r'$ is not always singleton. Alternatively, we choose a more general form where $\{\sigma\}$ should be $\projectTo{r}{\Code{FV}(P)}$, which at least contains all variables needed by $P$.
\end{definition}

\paragraph{Persistent type} Formally, a persistent type $\Gamma \vdash e : \ctpers \tau$ indicates that the typing of $e$ does not rely on the capability from either the bindings in type context $\Gamma$ or \emph{the reduction of the term $e$}. Consider the following three functions:
\begin{align*}
  &z{:}\urt{\Unit}{\top} \vdash \zzlam{x}{z} : \ort{\Unit}{\top} \ctsarr \urt{\Nat}{\top}
  \\&\emptyset \vdash \zzlam{x}{\Code{nat\_gen()}} : \ort{\Unit}{\top} \ctsarr \urt{\Nat}{\top}
  \\&\emptyset \vdash \zzlam{x}{0} \oplus \zzlam{x}{\Code{positive\_gen()}} : \ort{\Unit}{\top} \ctsarr \urt{\Nat}{\top}
\end{align*}\noindent
All of them can have the normal function types, while only the second one can be persistent. The first one is obviously not persistent since it depends on the binding $z$ in the type context, while the third one is also not persistent since its reachability guarantee is \emph{dependent on the reduction of the term}. The following type judgement fails in our call-by-value semantics:
\begin{align*}
  \emptyset \vdash \zlet{f}{\zzlam{x}{0} \oplus \zzlam{x}{\Code{positive\_gen()}}}{\zif{f\;() = 0}{f\;()}{0}} : \urt{\Nat}{\top} \quad \judgfail
\end{align*}\noindent
This term can only reach the value $0$ since $f\;()$ can either reduce to $0$ or to any positive number. In other words, the two applications of $f$ are indeed dependent, and $f$ is not persistent.

\begin{definition}[Persistent type] The type denotation of a persistent type $\ctpers \tau$ is defined as:
  \begin{align*}
    \denot{\ctpers \tau}_\Sigma\;e \doteq \forall x.\denot{\mstep{e}{x}} \chimpl \chpers (\denot{\tau}_{\Sigma,x{:}\eraserf{\tau}}\;x) \quad (\text{fresh $x$})
  \end{align*}\noindent 
Note that defining the type denotation as $\chpers (\denot{\tau}_{\Sigma}\;e)$ would also allow wrong inhabitants such as $\zzlam{x}{0} \oplus \zzlam{x}{\Code{positive\_gen()}}$. To ensure that $e$ is also single-valued, we introduce a universally quantified fresh variable $x$ that ranges over all possible reduction results of $e$; the persistent modality then ensures only that $x$ is single-valued.
\end{definition}

\subsection{Type Denotation for Nondeterminism}\label{subsec:nondet-denotation}

Nondeterminism breaks the single-valued assumption of the deterministic
language in two ways that require adjustments to the type denotation.

\paragraph{Function Types} The current rules for let binding are inconsistent with the type denotation in the nondeterministic setting. Consider the following judgement:
\begin{align*}
  \emptyset \vdash \zlet{x}{\Code{bool\_gen()}}{\zlam{y}{\Bool}{y = x}} : y{:}\urt{\Bool}{\top} \ctsarr \urt{\Bool}{\top}
\end{align*}\noindent
The current type denotation assumes that the parameter type $y$ cannot depend on the choice of $x$, which is hidden inside the generator invocation within the term; thus this function should return all boolean values. However, the let-expression rule (e.g., \textsc{T-Let}) exposes the hidden capability within the nondeterministic term:
\begin{align*}
  x{:}\urt{\Bool}{\top} \vdash \zlam{y}{\Bool}{y = x} : y{:}\urt{\Bool}{\top} \ctsarr \urt{\Bool}{\top}
\end{align*}\noindent
where $y$ can be chosen equal to $x$, so the function can only return $\true$.
The issue here is that a nondeterministic term exposes more capabilities into the environment during evaluation, which is not accounted for by the current type denotation.

\paragraph{Sum Types}
A similar issue arises for sum types.  The judgement:
\begin{align*}
  \emptyset \vdash \Code{bool\_gen()} :
  \ort{\Bool}{\vnu = \true} \ctoplus \ort{\Bool}{\vnu = \false}
\end{align*}
should hold, since $\Code{bool\_gen()}$ nondeterministically produces
either $\true$ or $\false$.  However, the original denotation
$\denot{\tau_1 \ctoplus \tau_2}\;e \doteq \denot{\tau_1}\;e \choplus
\denot{\tau_2}\;e$ requires the \emph{context} to be split across the
two branches, not the term itself.  Under the empty context, there is
no capability to split, so the judgement fails despite being
semantically valid.

\paragraph{Generalized Denotation}
Both issues stem from the same root cause: the original denotation
evaluates $e$ once and reasons about the result, but a
nondeterministic $e$ can reduce to different values in different
executions.  The fix is to universally quantify over all possible
reduction results of $e$, lifting $e$'s nondeterminism into the
type context:
\begin{align*}
  \denot{x{:}\tau_x \ctsarr \tau}_\Sigma\;e
    &\doteq \forall f.\; \denot{f \doteq e}_\Sigma \chimpl
             \denot{\tau_x}_\Sigma\;x \chimpl
             \denot{\tau}_{\Sigma,f{:}\eraserf{\tau_x \ctsarr \tau}, x{:}\eraserf{\tau_x}}\;(f\;x)
  \\\denot{x{:}\tau_x \ctwand \tau}_\Sigma\;e
    &\doteq \forall f.\; \denot{f \doteq e}_\Sigma \chimpl
             \denot{\tau_x}_\Sigma\;x \chwand
             \denot{\tau}_{\Sigma,f{:}\eraserf{\tau_x\ctwand \tau}, x{:}\eraserf{\tau_x}}\;(f\;x)
  \\\denot{\tau_1 \ctoplus \tau_2}_\Sigma\;e
    &\doteq \forall x.\; \denot{x \doteq e}_\Sigma \chimpl
             \denot{\tau_1}_{\Sigma,x{:}\eraserf{\tau_1}}\;x \choplus \denot{\tau_2}_{\Sigma,x{:}\eraserf{\tau_2}}\;x
\end{align*}
In each case, a universally quantified fresh variable ($f$ or $x$)
ranges over all possible reduction results of $e$, and the original
denotation is applied to that variable.  In the deterministic case,
$e$ has a unique reduction result and the quantifier is trivially
eliminated, so the generalized denotation coincides with the original.

For the function-type issue, the generalized denotation successfully captures the newly produced capabilities induced by the nondeterministic term. Thus the previous function-type judgement fails under the generalized denotation:
\begin{align*}
  &\denot{y{:}\urt{\Bool}{\top} \ctsarr \urt{\Bool}{\top}}\;
    (\zlet{x}{\Code{bool\_gen()}}{\zlam{y}{\Bool}{y = x}})
  \\=\;& \forall f.\; \Code{Atom}(f = (\zlam{y}{\Bool}{y = \true}) \lor
          f = (\zlam{y}{\Bool}{y = \false})) \chimpl
          \\& \quad \forall y. \Code{Atom}(y = \true \lor y = \false) \chimpl  \angelic\,\Code{Atom}(f\;y = \true \lor f\;y = \false)
  \quad\judgfail
\end{align*}
For the sum-type issue, the generalized denotation lifts
$\Code{bool\_gen()}$'s nondeterminism into the context via the fresh
variable $x$, allowing the sum operator to partition it across the
two branches as usual:
\begin{align*}
  &\denot{\ort{\Bool}{\vnu = \true} \ctoplus
          \ort{\Bool}{\vnu = \false}}\;\Code{bool\_gen()}
  \\=\;& \forall x.\; \Code{Atom}(x = \true \lor x = \false) \chimpl
          \demonic\,\Code{Atom}(\vnu = \true) \choplus
          \demonic\,\Code{Atom}(\vnu = \false)
  \quad\judgok
\end{align*}

\subsection{Case Studies}\label{subsec:case-studies}

We illustrate the expressiveness of context types with two further
case studies beyond the BST generator of Example~\ref{ex:coverage}:
a reverse reachability property for list reversal, and a disprove
reasoning task for a merge function.

\subsubsection{Coverage type}\label{subsubsec:coverage}

\begin{example}[Coverage Types]\label{ex:coverage}
  The coverage type system of~\citet{CoverageType} verifies that a
  test generator produces every desirable input.  Consider the
  following BST generator:
\begin{minted}[xleftmargin=5pt, numbersep=4pt, linenos = true, fontsize = \footnotesize, escapeinside=??]{OCaml}    
let rec bst_gen (r: nat) (lo: nat)
                (range: nat -> nat -> nat): nat tree =
  if r <= 0 then Leaf
  else Leaf ?$\oplus$?
    let r' = range (1, r) in
    let left  = bst_gen (r' - 1) lo range in
    let right = bst_gen (r - r') (lo + r') range in
    Node r' left right
\end{minted}
  The following specification asserts that for any valid range oracle,
  \texttt{bst\_gen} produces every BST whose keys fall in $(\Code{lo},
  \Code{lo}+r]$:
  \begin{align*}
    &\Code{val bst\_gen} :
      r{:}\ort{\Nat}{\top} \ctsarr lo{:}\ort{\Nat}{\top} \ctsarr
    \\&\quad\quad
      \Code{range}{:}\underbrace{(x{:}\ort{\Nat}{\top} \ctsarr
        y{:}\ort{\Nat}{x < \vnu} \ctsarr
        \urt{\Nat}{x < \vnu < y})}_{\text{demonic parameters; angelic result}}
      \ctwand
    \\&\quad\quad
      \urt{\Nat\;\Code{tree}}{\Code{bst}(\vnu) \land
        \forall x.\, \Code{mem}(\vnu, x) \iff
        (\Code{lo}+1 \leq x \leq \Code{lo}+r)}
  \end{align*}
  The range oracle has demonic parameter types ($\ort{\Nat}{\top}$ and
  $\ort{\Nat}{x < \vnu}$) and an angelic result type ($\urt{\Nat}{x <
    \vnu < y}$).  The demonic parameters mean that successive calls to
  \texttt{range} with different arguments can be typed with
  independent result capabilities — the oracle's output at each
  recursive call is disjoint from its outputs at other calls.  This is
  precisely the resource duplication property established above: the
  demonic parameter types make the oracle's capability persistent in
  negative position, enabling the recursive calls to accumulate
  independent angelic capabilities for the left and right subtrees.
  The use of the disjoint function type, $\ctwand$, on the oracle
  ensures it is disjoint from the size $r$ and lower-bound
  $\Code{lo}$, so the oracle's choices do not depend on those
  parameters.  The angelic result type then asserts that every valid
  BST is reachable by some sequence of oracle choices.
\end{example}

\subsubsection{Reverse Reachability}\label{subsubsec:reverse}

The standard list reversal function admits two distinct reachability
specifications:
\begin{lstlisting}
let rec reverse (l: list nat) : list nat =
  match l with
  | []      -> []
  | x :: xs -> reverse xs ++ [x]
\end{lstlisting}
\begin{align*}
  &\Code{val reverse} :
    l{:}\ort{\Nat\;\Code{list}}{\top} \ctsarr
    \urt{\Nat\;\Code{list}}{\Code{rev}(\vnu, l)}
  \\&\Code{val reverse} :
    l{:}\urt{\Nat\;\Code{list}}{\Code{sortedIncr}(\vnu)} \ctsarr
    \urt{\Nat\;\Code{list}}{\Code{sortedDecr}(\vnu)}
\end{align*}
The first uses a demonic input type: for any list $l$,
\texttt{reverse} produces its reversal.  The second uses an angelic
input type: when $l$ ranges over all increasingly-sorted lists, the
output ranges over all decreasingly-sorted lists.  The second
specification is strictly stronger, demonstrating that context types
can express reachability properties that depend on the full range of
the input.

\subsubsection{Disprove Reasoning}\label{subsubsec:disprove}

The following function merges two lists element-by-element:
\begin{lstlisting}
let rec merge (l1: elem list) (l2: elem list) : elem list =
  match l1, l2 with
  | l1, []          -> l1
  | [], l2          -> l2
  | h1::t1, h2::t2 ->
      let t = merge t1 t2 in
      if h1 < h2 then h1 :: h2 :: t
      else if h1 > h2 then h2 :: h1 :: t
      else h1 :: t
\end{lstlisting}
Context types can express both positive reachability specifications
and \emph{disprove} reasoning — showing that a function cannot always
produce a sorted output.  The following two specifications hold:
\begin{align*}
  &\Code{val merge} :
    l_1{:}\ort{\Nat\;\Code{list}}{\top} \ctsarr
    l_2{:}\urt{\Nat\;\Code{list}}{\top} \ctwand
    (\ort{\Nat\;\Code{list}}{\neg\Code{sortedIncr}(\vnu)}
      \ctoplus \ort{\Nat\;\Code{list}}{\Code{sortedIncr}(\vnu)})
  \\&\Code{val merge} :
    l_1{:}\urt{\Nat\;\Code{list}}{\top} \ctwand
    l_2{:}\urt{\Nat\;\Code{list}}{\top} \ctwand
    (\ort{\Nat\;\Code{list}}{\neg\Code{sortedIncr}(\vnu)}
      \ctoplus \ort{\Nat\;\Code{list}}{\top})
\end{align*}
The first says: when $l_2$ ranges over all lists, the output either
fails to be sorted or is sorted.  The second strengthens this: when
both inputs range over all lists, the output can fail to be sorted.
The following specification correctly fails to type-check, confirming
that \texttt{merge} cannot be specified as always producing a sorted
output:
\begin{align*}
  &\Code{val merge} :
    l_1{:}\ort{\Nat\;\Code{list}}{\top} \ctsarr
    l_2{:}\ort{\Nat\;\Code{list}}{\top} \ctsarr
    \ort{\Nat\;\Code{list}}{\Code{sortedIncr}(\vnu)}
  \quad\judgfail
\end{align*}
This failed judgement serves as a reachability-based witness to the
function's incorrectness: there exist inputs for which
\texttt{merge} does not produce a sorted list.
