\section{Supporting Nondeterminism}
\label{sec:extension}

The \langname{} language of Section~\ref{sec:context-typing} is
deterministic: every term reduces to (at most) one value under any
particular environment. This section considers a calculus,
\langNDname{}, that relaxes this restriction, enabling us to, e.g.,
type programs like \ref{ex:DisjCon} from
\autoref{sec:overview}. Concretely, \langNDname{} extends \langname{}
with nondeterministic \emph{generators} that can produce multiple
values of basic type across different executions:
\begin{align*}
  \primop ::= \quad & \ldots ~|~ \Code{gen}_b ~|~ \ldots
\end{align*}
Unsurprisingly, the wrinkle introduced by nondeterminism is related to
dependence: whereas the reachability properties of a \langname{} term
were fundamentally tied with the environment under which it is
evaluated, those of a \langNDname{} term may (or may not) be
\emph{independent} of the surrounding context.

Dealing with this new possibility raises two technical
challenges. First, nondeterministic functions like generators can be
invoked multiple times, producing independent capabilities at each
call; our extended type system must account for this \emph{resource
  duplication}.  Second, we need to generalize our type denotation to
accommodate terms whose reachability guarantees are not completely
determined by their evaluation environment.

% (\S\ref{subsec:nondet-denotation}).  We conclude with case studies
% (\S\ref{subsec:case-studies}) demonstrating the expressiveness of
% the resulting system.

% The nondeterministic choice $e_1 \oplus e_2$ is then expressible as
% $\zif{\Code{gen}_\Code{bool}\Code{()}}{e_1}{e_2}$.  These generators serve as the
% ``impure external environments'' introduced informally in
% Section~\ref{sec:context-typing}: the environment $\Envir_x(e) \doteq
% \zlet{x}{\Code{nat\_gen()}}{e}$ is now a first-class term that
% introduces a nondeterministic binding for $x$ whose capability is
% \emph{independent} of the surrounding context at each invocation.

\subsection{Resource Duplication and Persistence} %
\label{subsec:duplication}

In \langname{}, the reachability properties of a term are entangled
with its surrounding context: evaluation does not introduce new
capabilities. For example, every inhabitant of
$\ort{\Unit}{\top} \ctsarr \urt{\Nat}{\vnu > 0}$ in \langname{} must
capture some variable with an angelic binding that provides the
witnesses needed to construct every positive number:
\begin{align*}
  z{:}\urt{\Nat}{\top} \vdash \zzlam{x}{z + 1} : \ort{\Unit}{\top} \ctsarr
  \urt{\Nat}{0 < \vnu}
\end{align*}\noindent
Two successive calls to a \langname{} function with this signature
\emph{always} yield entangled results, even if the calls have
independent arguments:
\begin{align}
  z{:}\urt{\Nat}{\top} \vdash
  \zlet{f}{\zzlam{x}{z}}{\zlet{y_1}{f\;()}{\zlet{y_2}{f\;()}{y_1 -
  y_2}}} : \urt{\Nat}{\top} \quad \judgfail
  \tag{$\Code{e_{ND}}$}\label{ex:nondet}
\end{align}\noindent
This fact is encoded in the typing rules for let-expressions,
\textsc{T-Let} and \textsc{T-LetI}, which jointly enforce an
``entangled xor consumed'' policy. The rule \textsc{T-Let} allows the
function $f$ to be used in both the bound expression and the body of a
\Code{let}-expression, but the bound expression will be entangled with
$f$. Thus, if we tried to use \textsc{T-Let} to establish
\ref{ex:nondet}, we would get stuck when trying to type
$y_1 - y_2$:

\begin{center}
\small
  \begin{prooftree}
    \hypo{\ldots \ctcomma y_1{:}\urt{\Nat}{\top} \ctcomma
    y_2{:}\urt{\Nat}{\top} \vdash y_1 - y_2 : \urt{\Nat}{\top} \quad \judgfail}
  \infer1[\textsc{T-Let}]{\emptyset \ctcomma z{:}\urt{\Nat}{\top} \ctcomma f{:}\ort{\Unit}{\top}
    \ctsarr \urt{\Nat}{\top} \ctcomma y_1{:}\urt{\Nat}{\top} \vdash
    \zlet{y_2}{f\;()}{y_1 - y_2} : \tau}
\infer1[\textsc{T-Let}]{\emptyset \ctcomma z{:}\urt{\Nat}{\top}
  \ctcomma f{:}\ort{\Unit}{\top} \ctsarr \urt{\Nat}{\top} \vdash
  \zlet{y_1}{f\;()}{\zlet{y_2}{f\;()}{y_1 - y_2}} : \tau}
\end{prooftree}
\end{center} \smallskip

\noindent On the other hand, \textsc{T-LetI} produces an independent
result, but prevents $f$ from reappearing in the body of a
\Code{let}-expression, leaving us unable to type the second
\Code{let}-expression in \ref{ex:nondet}: \smallskip

\begin{center}
\small
  \begin{prooftree}
    \hypo{\emptyset \ctstar y_1{:}\urt{\Nat}{\top} \vdash
      \zlet{y_2}{f\;()}{y_1 - y_2} : \tau \quad \judgfail}
  \infer1[\textsc{T-LetI}]{\emptyset \ctstar (z{:}\urt{\Nat}{\top}
    \ctcomma f{:}\ort{\Unit}{\top} \ctsarr \urt{\Nat}{\top}) \vdash
    \zlet{y_1}{f\;()}{\zlet{y_2}{f\;()}{y_1 - y_2}} : \tau}
  \end{prooftree}
\end{center} \smallskip

Successive applications of a nondeterministic generator like
$\Code{gen_{nat}\; ()}$, however, can produce genuinely independent
values. In order to accurately capture the behavior of such functions,
\langNDname{} introduces a \emph{persistent} type operator, $\ctpers$:
\begin{align*}
  \Code{gen_{nat}} : \ctpers (\ort{\Unit}{\top} \ctsarr \urt{\Nat}{\top})
\end{align*}
Persistence allows a binding in the type context to be duplicated,
e.g.,
\begin{align*}
  x{:}\ctpers \tau \; \ctsubseteq{\{x\}} \; x{:}\ctpers \tau \ctstar x{:}\ctpers \tau
\end{align*}\noindent
When combined with \textsc{T-LetI}, a persistent resource can be
freely duplicated across multiple function applications without
entangling their results.
% BD: A careful reader will notice that this seems to contradict the
% assertion that "variable bindings in a (well-formed) type context
% [need] to be unique"  (Line 450)

\paragraph{Duplicable resource and persistent modality} In other
BI-logics that support the persistent modality~\cite{Iris}, its
definition is typically based on a \emph{duplicable} resource with
respect to the product operation of the underlying BI-algebra. A
duplicable resource is idempotent with respect to the product
operation, i.e., $r \times r = r$, which is what allows us to
duplicate generator types in the type context.  In our setting,
$r \times r = r$ only holds when $r$ is a \emph{singleton} capability
containing exactly one environment. Following this idea, we define the
persistent modality to hold on formulas that do not depend on a
capability providing different choices:

\begin{definition}[Persistent Modality] We introduce a persistence
  modality $\chpers P$ to our context logic such that:
  \begin{align*}
    r \models \chpers P \text{ iff } \exists \sigma. \{\sigma\} \sqsubseteq r \land \{\sigma\} \models P
  \end{align*}\noindent
  Note that we do not require $r$ to be singleton over its entire
  variable domain, as doing so would violate Kripke monotonicity: if a
  capability $r$ is singleton, then for $r'$ such that
  $r' \sqsubseteq r$, $r'$ is not always singleton. Instead, we impose
  the more relaxed requirement that there is some
  $\{\sigma\} \sqsubseteq r$, which need only include the variables
  relevant to $P$, i.e. $\projectTo{r}{\Code{FV}(P)}$.
\end{definition}

\paragraph{Persistent type} The persistent type
$\Gamma \vdash e : \ctpers \tau$ indicates that the typing of $e$ does
not rely on the capabilities provided by either the bindings in the
type context $\Gamma$ or \emph{the reduction of the term $e$}. To
illustrate, consider the following three functions:\footnote{The
  nondeterministic choice operator $e_1 \oplus e_2$ is syntactic sugar
  for $\zif{\Code{gen_{bool} \;()}}{e_1}{e_2}$.}
\begin{align*}
  &z{:}\urt{\Nat}{\top} \vdash \zzlam{x}{z} : \ort{\Nat}{\top}
    \ctsarr \urt{\Nat}{\top} \judgok \\
  &\emptyset \vdash \zzlam{x}{\Code{gen_{nat} \;()}} : \ort{\Unit}{\top}
    \ctsarr \urt{\Nat}{\top}  \judgok \\
  &\emptyset \vdash \zzlam{x}{0} \oplus
    \zzlam{x}{\Code{gen_{nat} \;() + 1}} : \ort{\Unit}{\top} \ctsarr
    \urt{\Nat}{\top}  \judgok
\end{align*}\noindent
Of these three terms, only the second can also be assigned a
persistent modality. The first is obviously not persistent since it
depends on the binding $z$ in the type context, while the reachability
properties of the third \emph{depend on how it is reduced}. To see
how, consider the following type judgment, which is semantically
invalid under \langNDname{}'s call-by-value reduction strategy:
\begin{align*}
  \emptyset \vdash \zlet{f}{\zzlam{x}{0} \oplus
  \zzlam{x}{\Code{gen_{nat} \;() + 1}}}{\zif{f\;() = 0}{f\;()}{0}} : \urt{\Nat}{\top} \quad \judgfail
\end{align*}\noindent
This term can only reach the value $0$: since $f$ is bound to either
$\zzlam{x}{0}$ or $\zzlam{x}{\Code{gen_{nat} \;() + 1}}$ when the
first let expression is reduced, two applications of $f\;()$ will
either both reduce to $0$ or some positive number. In other words, the
two applications of $f$ are indeed dependent, and $f$ is not
persistent.

\begin{definition}[Persistent Type Denotation]
  The denotation of $\ctpers \tau$ is:
  \begin{align*}
    \denot{\ctpers \tau}_\Sigma\;e \doteq \forall x.\denot{\mstep{e}{x}} \chimpl \chpers (\denot{\tau}_{\Sigma,x{:}\eraserf{\tau}}\;x) \quad (\text{fresh $x$})
  \end{align*}\noindent
  Note that if we had instead defined the type denotation as
  $\chpers (\denot{\tau}_{\Sigma}\;e)$, it would (incorrectly) contain
  inhabitants like
  $\zzlam{x}{0} \oplus \zzlam{x}{\Code{gen_{nat} \;() + 1}}$. To
  ensure that $e$ is single-valued, we introduce a universally
  quantified fresh variable $x$ that ranges over all possible
  reduction results of $e$; the persistent modality then ensures both
  the context and this $x$ are single-valued.
\end{definition}

\paragraph{Typing rules for $\ctpers$} The typing rules for persistent
types are:
\begin{align*}
\mprooftr{20mm}
      {$\Gamma \vdash e : \ctpers \tau$}
      {T-PersE}
      {$\Gamma \vdash e : \tau$}
\quad
  \mprooftr{40mm}
      {$\Gamma \vdash v : \tau$ \quad $\denot{\Gamma} \models \chpers \denot{\Gamma}$}
      {T-PersI}
      {$\Gamma \vdash v : \ctpers \tau$}
\end{align*}\noindent
The elimination rule for persistent types \textsc{T-PersE} is
derivable from \textsc{T-Sub} and the fact
$\Gamma \vdash \ctpers \tau <: \tau$ always holds.  Its introduction
rule \textsc{T-PersI} allows us to conclude that a value is a
persistent term of type $\ctpers \tau$ if we can establish that it has
type $\tau$ using a type context comprised only of persistent facts,
i.e., $\denot{\Gamma} \models \chpers \denot{\Gamma}$.

%% Note that
%% although \textsc{T-PersI} can only be directly applied to values, $v$,
%% it is always possible to let-expand a persistent expression into a
%% form which the rule can be applied to, e.g., by transforming
%% $(\zzlam{y}{\Code{gen_{nat} \;()}}) 3$ to
%% $\zlet{x}{(\zzlam{y}{\Code{gen_{nat} \;()}}) 3}{x}$.
%\BD{This example is not working for me: as is, the term is not
%persistent. If we instead eta expand the body, we end up with a
%term... We either need to come up with a better example or drop this
%discussion entirely.}


\subsection{Type Denotation for Nondeterminism} %
\label{subsec:nondet-denotation}

When reasoning about reachability properties of \langname{} terms, all
the choices available to us came from angelic refinements in the
typing context. In contrast, nondeterministic \langNDname{} terms can
themselves introduce choices that are not fully captured by our
previous type denotation. This prevents us from applying semantic
subtyping to verify properties of terms which should intuitively hold.
To illustrate, consider the following typing derivation: \smallskip
\begin{center}
\small
\begin{prooftree}
  \hypo{\emptyset \vdash \Code{gen_{bool} \;()}  : \urt{\Bool}{\top}}
  \hypo{\emptyset \vdash \urt{\Bool}{\top} \subty \ort{\Bool}{\vnu = \true} \ctoplus \ort{\Bool}{\vnu = \false}}
  \infer2[T-CtxSub]{\emptyset \vdash \Code{gen_{bool} \;()} : \ort{\Bool}{\vnu =
      \true} \ctoplus \ort{\Bool}{\vnu = \false}}
\end{prooftree}
\end{center}\smallskip

Intuitively, the derivation should be valid: $\Code{gen_{bool} \;()}$
exhibits two control flow paths, reaching \Code{true} on one and
\Code{false} on the other. Unfortunately, this conclusion does not
hold \emph{semantically} under our original denotation, i.e.,
$\not \models~ \denot{\ort{\Bool}{\vnu = \true} \ctoplus
  \ort{\Bool}{\vnu = \false}}~ \Code{gen_{bool} \;()} $. The issue
here is that the original denotation of $\choplus$ only splits the
capabilities of the \emph{context} between the two branches, not the
term itself --- for our target claim to hold, $\Code{gen_{bool} \;()}$
could only evaluate to \Code{true} under some partitioning of the
empty context, which clearly never happens!

A similar issue arises with the original denotation of function
types. The fix to both issues is to ``lift'' the extra capability
provided by a nondeterministic term into the capability context:
\begin{align*}
  \denot{x{:}\tau_x \ctsarr \tau}_\Sigma\;e
    &\doteq \forall f.\; \denot{\mstep{e}{f}}_\Sigma \chimpl
             \denot{\tau_x}_\Sigma\;x \chimpl
             \denot{\tau}_{\Sigma,f{:}\eraserf{\tau_x \ctsarr \tau}, x{:}\eraserf{\tau_x}}\;(f\;x)
  \\\denot{x{:}\tau_x \ctwand \tau}_\Sigma\;e
    &\doteq \forall f.\; \denot{\mstep{e}{f}}_\Sigma \chimpl
             \denot{\tau_x}_\Sigma\;x \chwand
             \denot{\tau}_{\Sigma,f{:}\eraserf{\tau_x\ctwand \tau}, x{:}\eraserf{\tau_x}}\;(f\;x)
  \\\denot{\tau_1 \ctoplus \tau_2}_\Sigma\;e
    &\doteq \forall x.\; \denot{\mstep{e}{x}}_\Sigma \chimpl
             \denot{\tau_1}_{\Sigma,x{:}\eraserf{\tau_1}}\;x \choplus \denot{\tau_2}_{\Sigma,x{:}\eraserf{\tau_2}}\;x
\end{align*}
In each case, a universally quantified fresh variable ($f$ or $x$)
ranges over all possible reduction results of $e$, and the original
denotation is applied to that variable.  For deterministic terms,
which have a unique reduction result, the quantifier is trivially
eliminated, so the generalized denotation coincides with the original
when $e$ is deterministic.

This new denotation allows us to lift $\Code{gen_{bool} \;()}$'s
nondeterminism into the context via the fresh variable $x$, allowing
the sum operator to partition it across the two branches as before:
\begin{align*}
  &\denot{\ort{\Bool}{\vnu = \true} \ctoplus
          \ort{\Bool}{\vnu = \false}}\;\Code{gen_{bool} \;()}
  \\=\;& \forall x.\; \Code{Atom}(x = \true \lor x = \false) \chimpl
          \demonic\,\Code{Atom}(x = \true) \choplus
          \demonic\,\Code{Atom}(x = \false)
  \quad\judgok
\end{align*}
This richer denotation also allows us to discharge the semantic
subtyping assumption in the previous typing derivation, allowing us to
verify that $\Code{gen_{bool} \;()}$ meets this intuitive
specification. Finally, we can show that \langNDname{}'s extended type
system is sound with respect to this generalized type denotation.

\begin{theorem}[\langNDname{} Type Soundness]%
  \label{thm:nondet-soundness}
  For a well-typed \langNDname{} term $\emptyset \vdash e : \tau$ and
  fresh variable $x$,
  $\{[x \mapsto v] \mid \mstep{e}{v}\} \models \denot{\tau}\;x$.
\end{theorem}


% \paragraph{Function Types} The current rules for let binding are inconsistent with the type denotation in the nondeterministic setting. Consider the following judgement:
% \begin{align*}
%   \emptyset \vdash \zlet{x}{\Code{gen_{bool} ()}}{\zlam{y}{\Bool}{y = x}} : y{:}\urt{\Bool}{\top} \ctsarr \urt{\Bool}{\top}
% \end{align*}\noindent
% The current type denotation assumes that the parameter type $y$ cannot depend on the choice of $x$, which is hidden inside the generator invocation within the term; thus this function should return all boolean values. However, the let-expression rule (e.g., \textsc{T-Let}) exposes the hidden capability within the nondeterministic term:
% \begin{align*}
%   x{:}\urt{\Bool}{\top} \vdash \zlam{y}{\Bool}{y = x} : y{:}\urt{\Bool}{\top} \ctsarr \urt{\Bool}{\top}
% \end{align*}\noindent
% where $y$ can be chosen equal to $x$, so the function can only return $\true$.
% The issue here is that a nondeterministic term exposes more capabilities into the environment during evaluation, which is not accounted for by the current type denotation.

% For the function-type issue, the generalized denotation successfully captures the newly produced capabilities induced by the nondeterministic term. Thus the previous function-type judgement fails under the generalized denotation:
% \begin{align*}
%   &\denot{y{:}\urt{\Bool}{\top} \ctsarr \urt{\Bool}{\top}}\;
%     (\zlet{x}{\Code{bool\_gen()}}{\zlam{y}{\Bool}{y = x}})
%   \\=\;& \forall f.\; \Code{Atom}(f = (\zlam{y}{\Bool}{y = \true}) \lor
%           f = (\zlam{y}{\Bool}{y = \false})) \chimpl
%           \\& \quad \forall y. \Code{Atom}(y = \true \lor y = \false) \chimpl  \angelic\,\Code{Atom}(f\;y = \true \lor f\;y = \false)
%   \quad\judgfail
% \end{align*}
